---
title: "Week 4 Problem Set"
subtitle: "Orbital theory and Earth's past climate"
author: "Juliana Olsen-Valdez"
date: "Last knitted `r format(Sys.Date(), '%d %b %Y')`"
output:
  pdf_document:
    toc: yes
    toc_depth: '3'
  html_document:
    df_print: paged
    number_sections: yes
    toc: yes
    toc_float: yes
    toc_depth: 3
    code_folding: show

editor_options:
  chunk_output_type: inline
---
> Click into a code cell (the gray blocks below) to select or edit it. To run a selected code cell, hit `Ctrl`+`Shift`+`Enter` (Windows & Linux) / `Command`+`Shift`+`Enter` (MacOS). Make sure that each code cell runs successfully before you move on to the next one.

Load the libraries that we'll use in this Problem Set: 
```{r load libraries}
library(tidyverse)
library(dplyr)
library(ggplot2)
```

##Introduction
In this problem set, we'll be exploring orbital forcings and their effect on climate patterns, specifically with respect to glaciations in the last million years. Since the mid- to late-1800s, scientists have been studying the extent, frequency, and cause of glaciations on Earth. And though ideas and data has greatly expanded over that time, the role of the earth's orbital parameters (eccentricity, obliquity, and precession) has been at the forefront of this conversation since scientists first started questioning the climate and astronomical dynamics that lead to glacial and interglacial periods. 

##Learning Goals
By the end of this exercise, you will be able to:
1) Describe the three main types of orbital forcing, how they work, and at what time scale each occurs
2) Define the term insolation and provide its units
3) Use the "ggplot" function to create a plot of your own, using aesthetic parameters to add labels and titles
4) Use the "comment-out" symbol (#) to describe each step of your plotting code

#Prerequisites
Prior to completing this Problem Set, it is expected that students: 
-Know terms related to Earth like orbit, tilt, latitude, and glaciation
-Have a basic familiarity with the structure and syntax of code in R
-Know how to copy, cut, comment-out, and edit text and code in R
-Feel comfortable creating and manipulating simple data frames and tibbles in R

##Background and Defining Terms
Three astronomical (orbital) theories relate to the Earth: eccentricity, obliquity, and precession. Let's get right into it!

1) **Eccentricity** defines the *circularity of the earth's orbit* around the sun. When Earth has a more circular orbit, it's distance from the sun at any given time during a full orbit is not much changed. When Earth's orbit is more elliptical, like an oval egg, the Earth spends part of the year at the far ends of the orbit (farthest away from the sun) and part of the year closer to the sun. Eccentricity cycles are on the scale of 100 thousand years (kyr) and 400kyr. 

2) **Obliquity** defines the *tilt of Earth's axis*, which varies between about 22 and 24 degrees. When the obliquity is high (aka the tilt is closer to 25), higher latitudes can get more direct sunlight compared to a lower obliquity (where lower latitudes receive more direct sunlight). Obliquity occurs in 41kyr-cycles. 

3) **Precession** defines the *direction of the tilt axis* of the earth. The tilt direction is described relative to if its pointing towards the north star Vega or the north star Polaris, but the significance of precession is that it can influence which hemispheres receive the most sunlight depending on which hemisphere pole ends up "facing away" from the sun during a given time. Precession occurs in approximately 21kyr-cycles. 


These orbital pararmeters are significant when it comes to climate patterns because they govern the distance Earth is from the sun, what time of year that occurs, and how much direct sunlight Earth receives at different latitudes during different times of the year. The way we quantify the amount of sunlight the earth receives at a specific latitude at a specific time is through the term **insolation** $(W/m^2)$ (remember, that's an energy!). Insolation describes the amount of power or heat (in Watts) that the earth receives per square meter. Therefore, when insolation is lower, we assume that the tempearature and climate were cooler, and vice versa. 


Now that you're more familiar with the terminology behind these orbital forcings, let's just remind ourselves why all of this is important in understanding climate change patterns and glaciations. When you think about the earth today, we associate the tilt of Earth's axis as being towards the sun, at an angle of around 23.5 degrees, and the earth's orbit as being farthest away during summertime in the northern hemisphere. 

Earth's higher tilt towards the sun all lead to a warmer climate, but what if I told you there have been time's in Earth's past where in addition to that, the summer occurred when the earth was closest to the sun?! That would be a pretty hot summer - one that could begin to melt any snow or ice on the planet. And, what if I told you that there have been times where the earth had a high tilt away from the sun AND was farthest away from the sun in the summer? Those would be some pretty cool summers, right (and pretty chilly winters!)? 

The power of these orbital forcings, specifically in this last example (cool summers that keep ice/snow from melting) were recognized in the early to mid 1900s by a scientist named Milutin Milankovitch. He hypothesized that glaciation and ice ages start when snow and ice don't melt during cooler summers (not necessarily really cold winters). And today, we still use the behavior of orbital forcings at different times in the past to predict what climate looked like during those times! That's a powerful tool!

##Example
Now, we're going to use some real data that highlights insolation from different latitudes over time to get a visual on what these orbital trends look like! In the example, I'll also introduce how to create a function that can be used without having to complete numerous lines of command. 

We'll start by loading insolation data from Berger and Loutre from seven different latitudes over the last million years:
```{r load data}
insol91.jun <- read.csv("insol91.jun.csv", stringsAsFactors = FALSE) #This is a data file for the June solstice
insol91.dec <- read.csv("insol91.dec.csv", stringsAsFactors = FALSE) #This is a data file for the Decemeber solstice

insol91.jun_dec <- left_join(insol91.jun,insol91.dec, by = "INSOLATION")

insol91.jun_dec <- rename(insol91.jun_dec, Time = INSOLATION)
```



```{r plot summmer (June) solstice at 30N}
plot.jun_30N <- ggplot(insol91.jun_dec, aes(x = Time, y=X30NJune)) +
geom_point(size = 0.5, shape = 19, color = "red") +
geom_line(size = 0.5, color = "red") + 
scale_x_reverse() +
labs(y = "June Insolation at 30N (W/m^2)", x = "Time (kyr)")+
theme_bw() 

plot.jun_30N
```


```{r}
make_plot <- function(...) { #Making a plot as a function. Takes x axis as input
  ggplot2::ggplot(data=insol91.jun_dec) +
  #plot <- ggplot(insol91.jun_dec) +
  x.title <- paste("Time before present (kyr) for", toString(Time[i]),) +
  labs(x = x.title , y="Insolation")
  geom_point(aes(y = X30NJune, x = Time), size = 0.5, shape = 19, color = "red") +
  geom_line(size = 0.5, color = "red") + 
  scale_x_reverse() +
  theme_bw() +
  
  return(plot)
}

make_plot(y=X30NJune, x=Time)
```


##Questions
1) Choose ONE of the orbital paramters we went over in this problem set and summarize: 
a) what it is 
b) how it generally influences Earth's climate
c) the frequency of its cycle
d) an example of its forcing on climate change 

**Question 1 Solution** 
Eccentricity describes the circularity of Earth's orbit, where the orbit modulates between a circle and an ellipse. Eccentricity influences Earth's climate by changing the distance that the earth is from the sun during different seasons (winter solstice versus summer solstice). Eccentricity has two cycles; every 100kyrs and 400kyrs. Eccentricity can force climate by causing the summer to stay really cool if the earth's orbit is elliptical and experiences the summer solstice when it's farthest from the sun. This causes winter snow and ice to stay frozen and can create a positive feedback that leads to even colder winters and summers to follow, creating a glacial stage!

_______________________________________________________

2) Now that you've gained some familiarity with plotting using the "ggplot" function, use the Berger and Loutre data set provided above to plot and compare the insolation at two latitudes of your choice. Comment on how they differ in, 1) insolation magnitude $(W/m^2)$, 2) period (aka are the peaks off-set or do they line up?), and note any other similarities or differences you notice between the two. I expect your plot(s) to be well-labeled and your code to have comments for each line to describe what you did. 

**Question 2 Solution**
I'm going to compare the insolation $(W/m^2)$ between 90N in June and 90S in December.  

```{r plot summmer (June) at 90N and winter (Dec) solstice at 90S}
plot.jun_90N_dec_90S <- ggplot(insol91.jun_dec, aes(x = Time, y=X90NJune)) + #creating a new plot and providing the data frame and x, y values that I'll use
geom_line(aes(y=X90NJune), colour="red") + #label my 90NJune as red
geom_line(aes(y=X90SDec), colour="blue") + #label my 90SDec as blue
scale_x_continuous(limits = c(-1000,0), breaks=seq (-1000,0,100)) + #I add in unique axis tick values for Time (kyr)
labs(y = "June Insolation at 90N and Dec. Insolation at 90S (W/m^2)", x = "Time before 1991 (kyr)")+ #I add my x-axis and y-axis labels and a title
theme_bw() #I want my plot to show up in black and white theme

plot.jun_90N_dec_90S #I call my plot so that it displays
```
First off, because these latitudes are in different hemispheres, these should actually record similar insolation values (NH summer solstice happens at the same time as SH winter solstice). We see that! The magnitude of insolation for each latitude line up pretty well! It looks like there is a little offset in that the 90S December insolation slightly lags (less than a half a period because it's not out of phase) compared to the 90N June. Another note is that both latitudes have times when their insolation peaks are higher than the other. 
________________________________________________________

3) This problem set didn't really get into the scientific methodology behind measuring insolations or temperature-constrained parameters from the past to create an orbitally-influenced climate record. Do some of your own research to highlight one method that scientists use to either age or determine the temperature/climate from sediments and rocks (including ice!) during the last 60 million years. Write at least three sentences and cite any sources you use. 

**Question 3 Solution**
Climate scientists often use the ratios of elemental isotopes like oxygen $(^{18}O)$ and $(^{16}O)$ to estimate the air temperature recorded in ice cores over the hundreds of thousands of years. The ratio of heavy to light oxygen is thermodynamically influenced and can therefore provide an analog to the air temperature over time. Age can be determined based on layers in the ice that relate to patterns of seasonality. Together, ice cores can provide age and temperature information over time that, when plotted, can be investigated for orbital influences! (Source: https://earthobservatory.nasa.gov/features/Paleoclimatology_IceCores)

_________________________________________________________

##Sources
-https://www.ncdc.noaa.gov/paleo-search/study/5776
-https://earthobservatory.nasa.gov/features/Paleoclimatology_IceCores

