---
title: "Week 3 Problem Set"
subtitle: "Ocean Acidification"
author: "Juliana Olsen-Valdez"
date: "Last knitted `r format(Sys.Date(), '%d %b %Y')`"
output:
  pdf_document:
    toc: yes
    toc_depth: '3'
  html_document:
    df_print: paged
    number_sections: yes
    toc: yes
    toc_float: yes
    toc_depth: 3
    code_folding: show

editor_options:
  chunk_output_type: inline
---
> Click into a code cell (the gray blocks below) to select or edit it. To run a selected code cell, hit `Ctrl`+`Shift`+`Enter` (Windows & Linux) / `Command`+`Shift`+`Enter` (MacOS). Make sure that each code cell runs successfully before you move on to the next one.

Load the libraries that we'll use in this Problem Set: 
```{r load libraries}
library(dplyr)
library(ggplot2)
```

##Introduction
Ocean acidification is a hot topic when it comes to present day climate change and global warming. In this problem set, we'll explore the how and why behind ocean acidification by using chemical reactions to explain the process and apply pH and saturation state calculations through R coding methods.  

##Learning Goals
By the end of this exercise, you will be able to:
1) Explain what ocean acidification is and how it works
2) Derive the chemical reactions that relate to ocean acidification and carbonate dissolution
3) Use the tibble function to make a data frame with various values related to ocean acidification in seawater
4) Make calculations using your data frame values to compare the effects of ocean acidification on different seawater chemistries

#Prerequisites
Prior to completing this Problem Set, it is expected that students: 
-Have a basic understanding of general chemistry, specifically with terms like pH, dissolution, and saturation state (omega) 
-Have a basic familiarity with the structure and syntax of code in R
-Know how to copy, cut, comment-out, and edit text and code in R

##Background and Defining Terms
#! [Alt text] (/Documents/CU Boulder Classes/Computational Tools/2020-ps-jolsenvaldez/Tufte.png)
! [Alt text] (/Downloads/Ocean-Acidification_image.png)

Ocean acidification 



Let's start with some chemical reactions that break down how 

$$1)\ CO^2 + H_2O \longleftrightarrow\  H_2CO_3^{2-}$$


$$2)\ H_2CO_3^{2-} \longleftrightarrow\ H^+ + HCO_3^{-}$$


$$3)\ HCO_3^{-} \longleftrightarrow\  H^+ + CO_3^{2-}$$


$$4)\ HCO_3^{-} \longleftrightarrow\  H^+ + CO_3^{2-}$$


$$5)\ Ca^{2+} + CO_3^{2-} \longleftrightarrow\  CaCO_3$$

##Example
We're going to dig deep into the concepts and equations around ocean acidification. 


Let's plot these two data points just to practice using ggplot:
```{r Plot Mars temperatures with different albedos}
plot_Mars_albedo <- (Mars_albedos %>%
       ggplot() + 
       aes(x = albedo, y=Mars_TempC) + 
       scale_x_continuous(limits = c(0.15,0.70), breaks=seq (0.15,0.70,0.05)) +
       geom_point() +
       labs (title = "Temperature of Mars for two different albedos", x = "albedo", y= "Temperature (degrees Celsius)") +
       geom_line()) +
       theme_bw()
plot_Mars_albedo
```



##Questions
1) Shortly explain (2-3 sentences), in your own words, why ocean acidification occurs, and how it affects ocean ecosystems, particularly calcifying organism. Use the terms included in the Problem Set but keep it concise!

**Question 1 Solution**

Ocean acidification occurs when the atmospheric $CO^2$ increases and leads to a greater amount of carbonic acid $H_2CO_3^{2-}$ in seawater. Carbonic acid dissolves in seawater to create $[H^+]$ ions that bond to suspended carbonate $CO_3^{2-}$ and can dissolve carbonate that is bound in organism shells. Therefore, increased $[H^+]$ leads to lower pH (making seawater more acidic) and limits the $CO_3^{2-}$ available to calcifying organisms. 

_______________________________________________________

2) Though the pH of seawater remained mostly constant for millions of years before present, in the last hundred years, the pH of surface sea water has fallen from 8.2 to 8.1. Further, it's been hypothesized that the pH will drop another 0.4 (to a value of 7.7) by the year 2100 (Orr et al., 2005). Calculate the percent rise in acidity for the change in pH for past (8.2) to modern (8.1) and modern to future (7.7), keeping in mind that $[H^{+}] = 10^{pH}$. Make your calculation by creating a small data frame. Hint: if you're having trouble knowing where to start, go back to the code from the example in the Week 2 Problem Set. Once you get your percent changes, comment on the magnitude of the values; were they what you were expecting based on the 0.1 and 0.4 changes in pH? 

**Question 2 Solution:**

```{r pH change}
pH_change <- #name the data frame
  tibble (
    Description = c("past","modern", "future"), #give descriptions to the rows
    pH = c(8.2, 8.1, 7.7), #include past and modern pH values
    Hconc = (10^(-pH)), #calculate H+ concentration
    Hconc_diff = (lead(Hconc) / Hconc), #calculate the fractional difference in H+ concentrations and add a column with that value
    Hconc_diff_percent = (Hconc_diff-1)*100 #use the difference in H+ concentration to calculate the percent change in acidity
  )

pH_change #display the data frame
```

The percent change in acidity given a 0.1 change in pH is around 26%. That's a much greater percent change than it would seem based on the seemingly miniscule change in pH. Further, the 0.4 pH change that's predicted for 2100 has an approximately 150% increase in acidity, which is alarming.

________________________________________________________

3) Follow the steps below to practice using coding language to investigate the temperature of the earth given varying albedo values: 
   a) create a data frame which includes the relevant values and constants in the Stefan-Boltzmann law needed to calculate temperature.
   b) add a range of albedos (), Stefan-Boltzmann constant (SBc),  and a calculated temperature column for the data frame. 
   c) use the data frame to plot albedo versus temperature for the earth.
   d) comment on the trend of the ggplot you created. Do the different albedos change the Temperature by much? Is the trend linear?
Hint: look back to the example for all the code you'll need. Simply rename columns and data frames to reflect information for the earth instead of Mars!
 
_________________________________________________________

##Sources
-https://ocean.si.edu/ocean-life/invertebrates/ocean-acidification
-https://theotherco2problem.wordpress.com/what-happens-chemically/
-https://www.st.nmfs.noaa.gov/Assets/Nemo/documents/lessons/Lesson_3/Lesson_3-Teacher's_Guide.pdf