---
title: "Week 2 Problem Set"
subtitle: "The Stefan-Boltzmann Law and Albedo"
author: "Juliana Olsen-Valdez"
date: "Last knitted `r format(Sys.Date(), '%d %b %Y')`"
output:
  pdf_document:
    toc: yes
    toc_depth: '3'
  html_document:
    df_print: paged
    number_sections: yes
    toc: yes
    toc_float: yes
    toc_depth: 3
    code_folding: show

editor_options:
  chunk_output_type: inline
---
> Click into a code cell (the gray blocks below) to select or edit it. To run a selected code cell, hit `Ctrl`+`Shift`+`Enter` (Windows & Linux) / `Command`+`Shift`+`Enter` (MacOS). Make sure that each code cell runs successfully before you move on to the next one.

Load the libraries that we'll use in this Problem Set: 
```{r load libraries}
library(dplyr)
library(ggplot2)
```

##Introduction
Last week, we introduced the Stefan-Boltzmann Law and applied it to questions regarding astronomical energies and temperatures, specifically with respect to black bodies. This week, we'll build on the Stefan-Boltzmann Law to include terms like albedo and calculations related to climate. 

##Learning Goals
By the end of this exercise, you will be able to:
1) Describe what albedo is, how it relates to emissivity, and how it can be substituted into the Stefan-Boltzmann Law equation. 
2) Calculate the temperature of the earth given different values for albedo. 
3) Use a basic plotting function and data to visualize quantitative values related to albedo. 

##Background and Defining Terms
Let's first re-familiarize ourselves with the Stefan-Boltzmann Law:
$$ \begin{align} \frac {P}{A} = e \sigma T^4 \end{align}$$
Terms and Units: 
$P = Power\ (Watts)$  
$ A=Area\ (meters^2)$
$e=emissivity\ (unitless)$
$\sigma = Stephan-Boltzmann  \  constant\  (5.67x 10^{-8}\ W/m^2/T^4)$
$ T= Temperature\ (degrees\  Kelvin^4)$

The Stefan-Boltzmann equation creates a proportional relationship between energy ($\frac{W}{m^2}$) and temperature (kelvins) to the fourth power. compares an energy (power per unit area squared, with units in $\frac{W}{m^2}$) to temperature raised to the fourth (with units in $Kelvin^4$). 

We briefly touched on emissivity in the last problem set, but we'll be delving into that more this week. As a reminder, emissivity refers to the effectiveness of an object to emit an energy. For theoretical black bodies, which absorb all incident heat radiation, emissivity is equal to 1.

**Albedo** is a term that refers to the amount of light that a planet (or any surface) reflects away. This makes albedo kind of like an "opposite" to emissivity. Therefore, a body that perfectly reflects all light (like one covered in snow or ice) would have an albedo of 100% (usually written in decimal form to be 1.00). In contrast, a planet that has a dark surface (or a black body) would have an albedo of 0. And just like emissivity, albedo is unitless. 

We relate albedo to emissivity through the equation: 
$$ e = 1 - albedo $$
Let's stop and think about if this equation makes sense -> 
-We expect a blackbody to have an albedo of zero and an emissivity of 1 since all light is absorbed and none is reflected. e = 1-0, e = 1. Great! 
-We expect a perfect reflector to have an albedo of one and an emissivity of zero since no light is absorbed and all is reflected. e = 1-1, e = 0. Awesome! 

Now, let's rewrite the Stefan-Boltzmann Law to include albedo: 
$$ \frac {P}{A} = (1-albedo)\ \sigma T^4 $$
We'll be using albedo to calculate a more accurate temperature for the earth, plot how different albedos affect Earth's temperature, and think about Earth's climate. 

##Example
First, we'll start by using albedo to explore the difference in temperature on Mars if it were mostly covered in ice (some scientists think it was once an icy planet!) versus in its current red, sandy state. We'll be using a version of the Stefan-Boltzmann law where we'll solve for Temperature, but this week we'll substitute our $1-albedo$ term in for the emissivity. Here's what the initial equilibrium equation looks like: 
$$(1-albedo)*So*\pi R^2 = \sigma T^4*4\pi R^2$$
This equation says that the radiation received by a planet (now affected by the albedo of that planet) is equal to the radiation emitted by that planet. 

Here is the equation above solved for temperature (kelvins):
$$T = [(1-albedo)*So/4\sigma]^{1/4}$$
We'll use a solar constant for Mars of 591 $\frac {W}{m^2}$, the Stefan Boltzmann constant of $5.67x 10^{-8}$ and albedos of 0.70 for icy Mars and 0.15 for modern Mars.

This week, you'll be introduced to a new coding tool that makes it easy to define and plot data. We'll first use tibbles to create a data frame with columns and rows. 
```{r Create a data frame for the different albedos, constants, and values for Mars}
Mars_albedos <- #this is the name of our data frame
tibble(
  Description = c("Sandy Mars", "Icy Mars"), # Entering row names
  albedo = c(0.15, 0.70), # We're assigning our to values to the column "albedo"
  So = 591, # Solar constant value for Mars
  SBc = 5.67E-8, # Stefan-Boltzmann constant value
  Mars_TempK = (((1-albedo)*So)/(4*SBc))^(1/4), #calculate Temperature in Kelvin using the terms we defined in our data frame
  Mars_TempC = (Mars_TempK - 273.15) #calculate the Temperature in degrees Celsius
)
Mars_albedos$Mars_TempC_diff = lead(Mars_albedos$Mars_TempC) - Mars_albedos$Mars_TempC #calculate the difference in the TempC between Sandy and Icy Mars and then add a column
Mars_albedos #we call our named data frame and it displays below!
```
We found that the temperature of a Sandy, red Mars with albedo of 0.15 has a temperature of -56 degrees Celsius and the temperature of an Icy Mars with albedo of 0.70 is -106 degrees Celsius. The temperature difference between these two Mars' is about -50 degrees Celsius. First off, wow that's cold! But also, look at the difference in temperature just because we were comparing an ice-covered Mars versus modern Mars. 

Let's plot these two data points just to practice using ggplot:
```{r Plot Mars temperatures with different albedos}
plot_Mars_albedo <- (Mars_albedos %>%
       ggplot() + 
       aes(x = albedo, y=Mars_TempC) + 
       scale_x_continuous(limits = c(0.15,0.70), breaks=seq (0.15,0.70,0.05)) +
       geom_point() +
       labs (title = "Temperature of Mars for two different albedos", x = "albedo", y= "Temperature (degrees Celsius)") +
       geom_line()) +
       theme_bw()
plot_Mars_albedo
```

##Questions
1) Explain what albedo and heat energy has to do with the difference in the feeling of wearing a black shirt versus a white shirt on a bright, hot summer day. 

_______________________________________________________

2) Solve for the Temperature of the earth using its modern albedo of 0.37. Report your answer in degrees Celsius. Hint: you can do this with a variable and list like we did last week. 

________________________________________________________

3) Follow the steps below to practice using coding language to investigate the temperature of the earth given varying albedo values: 
   a) create a data frame which includes the relevant values and constants in the Stefan-Boltzmann law needed to calculate temperature.
   b) add a range of albedos (), Stefan-Boltzmann constant (SBc),  and a calculated temperature column for the data frame. 
   c) use the data frame to plot albedo versus temperature for the earth.
   d) comment on the trend of the ggplot you created. Do the different albedos change the Temperature by much? Is the trend linear?
Hint: look back to the example for all the code you'll need. Simply rename columns and data frames to reflect information for the earth instead of Mars!
 
_________________________________________________________

4) After completing questions 2 and 3, what do you think about the temperatures that you calculated? Do they seem too hot or too cold? If they weren't what you were expecting given the temerature of Earth today, what could be missing from the model that we considered in calculating the earth's temperature? 

_________________________________________________________
##Sources
-https://scied.ucar.edu/planetary-energy-balance-temperature-calculate#:~:text=Scientists%20use%20the%20term%20albedo,planet%20or%20surface%20reflects%20away.&text=In%20the%201800s%2C%20two%20scientists,called%20the%20Stefan%2DBoltzmann%20law.
-https://asterism.org/2019/04/12/how-bright-is-the-moon/#:~:text=From%20NASA's%20planetary%20sites%2C%20the,from%20the%20cloud%2Dcovered%20planet.&text=Earth's%20albedo%20is%200.37%3B%20Mars,Uranus%2C%200.51%3B%20Neptune%200.41.
-https://www.britannica.com/science/Stefan-Boltzmann-law
-https://www.deviantart.com/zackdowney/art/Stefan-Boltzmann-Law-407403124
-http://hyperphysics.phy-astr.gsu.edu/hbase/thermo/stefan.html
-https://www.youtube.com/watch?v=DQk04xqvVbU
-https://www.youtube.com/watch?v=JiWnMd9I8GE